Conceptualizing open health data platforms for low- and middle income countries

### 2024-04-15 index.tex
wip

Conceptually, the OpenHIE framework constitutes a framework for an open
digital platform. Openness for digital platforms refers to i) the use of
open boundary resources, that is, specifications for the various
healthcare specific workflows and information standards such as FHIR;
and ii) the use of open source components that are available as digital
public goods\citep{digitalpublicgoods}. If we are to use the OpenHIE
framework as an open data platform, we need to extend the standards,
technologies and architecture to include functionality for data sharing
and reuse. Distinguishing four types of data sharing
(Table~\ref{tbl-types-data-sharing}), the purpose of this paper is to
investigate how new standards and technologies that can establish
openness of HDPs can be integrated into the OpenHIE architecture
framework. The lack of detailed specifications and consensus of this
addition to OpenHIE currently stands in the way of development projects
that aim to establish HDPs in LMICs. In addition, we explicitly address
the requirement of downward scalability as we aim to implement health
data platforms in resource constrained setting of LMICs.

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1200}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3200}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5600}}@{}}
\caption{Types of data sharing and in relation to new standards and
technology enablers to create
openness.}\label{tbl-types-data-sharing}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Type of data sharing}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Technology enablers to create openness of health data platforms}
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Type of data sharing}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Technology enablers to create openness of health data platforms}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{1} & Data at the most granular (patient) level, which is
persisted and used to provide a longitudinal record. & \textbf{Bulk FHIR
API}: has by now been incorporated in all major FHIR implementations the
FHIR standard can be readily used to support data sharing to
patient-level data across a patient population
\citep{mandl2020push, jones2021landscape}. \\
\textbf{2} & Aggregated data, for example statistics for policy
evaluation and benchmarking & \textbf{SQL-on-FHIR specification}
\citep{sql-on-fhir}: provides a standardised approach to make FHIR work
well with familiar and efficient SQL engines that are most commonly used
in analytical workflows. Builds on FHIRPath \citep{hl72020fhirpath}
expressions in a logical structure to specify things like column names
and unnested items. Implementations of this approach are available or
forthcoming, including open source implementations such as Pathling
\citep{grimes2022pathling} and commercial offerings like Aidbox
\citep{aidbox}. \\
\textbf{3} & Data analytics modules, that provide access to work and
access the data. & \textbf{Federated learning (FL)}
\citep{rieke2020future} and \textbf{privacy- enhancing technologies
(PETs)} \citep{scheibner2021revolutionizing, jordan2022selecting}: new
paradigms that address the problem of data governance and privacy by
training algorithms collaboratively without exchanging the data itself.
Models can be trained on combined datasets and made available as open
source artifacts for decision support. Data analysts can use FL and PETs
to work with the data in a collaborative, decentralized fashion. \\
\textbf{4} & Trained models that have been derived from the data and can
be used stand-alone for decision support. & \textbf{ONNX} \citep{onnx}:
an open format built to represent machine learning models. \\
\end{longtable}

@@

In the paper, we present a design that extends the OpenHIE specification
to include the four types of data sharing mentioned above. Using the
full-STAC approach \citep{mehl2023fullstac} we combine open standards,
open technologies and open architectures into a coherent modular HDP
platform that can be configured and reused across a variety of
use-cases. We employ a formative, naturalistic evaluation to assess the
technical risk and efficacy of the design \citep{venable2016feds}. Given
that it is prohibitively expensive to evaluate the design a real-world
setting, we aim to minimize technological risks and maximize the
efficacy of the design by considering three examples of health data
platforms in LMICs, namely:

@@

@@

The main contributions of this paper are i) description of a framework
for the components of the Data \& Analysis Services that builds on
current best practices from the data engineering community into the
OpenHIE framework; and ii) evaluation of different implementations and
design options for type 1 and 2 data sharing within an extended OpenHIE
architecture. As such, it aims to inform future developments and
implementation of open digital health platforms in LMICs.

@@

@@

@@

@@

\begin{itemize}
\item
  Object of openness: what data-related resources should data platforms
  make available when opening up (e.g.~data, data products, datadriven
  insights, analytics modules)? Which user groups derive value from
  accessing data-related resources from data platforms (e.g.~data
  providers, data users, intermediaries, developers)?
\item
  Unit of analysis: what is platform-to-platform openness in the context
  of data platforms, given the expectation that different HDPs will
  emerge at various aggregation levels? How do we distinguish
  meta-platforms, forking, and platform interoperability?
\item
  Risk of openness: What are the novel (negative) implications of
  opening up data platforms? How can reflexivity in design help
  providers to resolve the negative implications of openness?
\item
  Answers/insights to above:

>>>>

Conceptually, the OpenHIE framework constitutes a framework for an open
digital platform. Openness for digital platforms refers to i) the use of
open boundary resources, that is, specifications for the various
healthcare specific workflows and information standards such as FHIR;
and ii) the use of open source components that are available as digital
public goods \citep{digitalpublicgoods}. If we are to use the OpenHIE
framework as an open data platform, we need to extend the standards,
technologies and architecture to include functionality for data sharing
and reuse. The lack of detailed specifications and consensus of this
addition to OpenHIE currently stands in the way of development projects
that aim to establish HDPs in LMICs. The purpose of this paper,
therefore, is to specificy how new standards and technologies can be
integrated into the OpenHIE architecture such that an open HDP can be
realized that supports the four types of data sharing as defined by de
Reuver at al.~(Table~\ref{tbl-types-data-sharing}).

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1200}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3200}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5600}}@{}}
\caption{Types of data sharing and in relation to new standards and
technology enablers to create openness. Taken from de Reuver et
al.~(2022).}\label{tbl-types-data-sharing}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Type of data sharing}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Technology enablers to create openness of health data platforms}
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Type of data sharing}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Technology enablers to create openness of health data platforms}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{1} & Data at the most granular (patient) level, which is
persisted and used to provide a longitudinal record. & \textbf{Bulk FHIR
API}: has by now been incorporated in all major FHIR implementations the
FHIR standard can be readily used to support data sharing to
patient-level data across a patient population
\citep{mandl2020push, jones2021landscape}. \\
\textbf{2} & Aggregated data, for example statistics for policy
evaluation and benchmarking & \textbf{SQL-on-FHIR specification}
\citep{sql-on-fhir}: provides a standardised approach to make FHIR work
well with familiar and efficient SQL engines that are most commonly used
in analytical workflows. Builds on FHIRPath \citep{hl72020fhirpath}
expressions in a logical structure to specify things like column names
and unnested items. Implementations of this approach are available or
forthcoming, including open source implementations such as Pathling
\citep{grimes2022pathling} and commercial offerings like Aidbox
\citep{aidbox}. \\
\textbf{3} & Data analytics modules, that provide access to work and
access the data. & \textbf{Federated learning (FL)}
\citep{rieke2020future} and \textbf{privacy- enhancing technologies
(PETs)} \citep{scheibner2021revolutionizing, jordan2022selecting}: new
paradigms that address the problem of data governance and privacy by
training algorithms collaboratively without exchanging the data itself.
Models can be trained on combined datasets and made available as open
source artifacts for decision support. Data analysts can use FL and PETs
to work with the data in a collaborative, decentralized fashion. \\
\textbf{4} & Trained models that have been derived from the data and can
be used stand-alone for decision support. & \textbf{ONNX} \citep{onnx}:
an open format built to represent machine learning models. \\
\end{longtable}

@@

In this paper, we present a design that extends the OpenHIE
specification to include the four types of data sharing mentioned above.
Using the full-STAC approach \citep{mehl2023fullstac} we combine open
standards, open technologies and open architectures into a coherent
modular HDP platform that can be configured and reused across a variety
of use-cases. We employ a formative, naturalistic evaluation to assess
the technical risk and efficacy of the design \citep{venable2016feds}.
Given that it is prohibitively expensive to evaluate the design a
real-world setting, we aim to minimize technological risks and maximize
the efficacy of the design by considering three examples of health data
platforms in LMICs, namely:

@@

These three real-world implementations are evaluated along the following
dimensions:

\begin{itemize}
\tightlist
\item
  What is the level of openness of the implementation, specifically in
  terms of the four types of data sharing and platform-to-platform
  opennenss?
\item
  How are the core maintenance functions of the HDP implemented, where
  we particularly focus on functions related to metadata management,
  data lineage and access control mechanisms?
\item
  Is the solution suitable for downward scalability, where we focus on
  the computer and storage requirements of the solution and assess the
  minimal requirements for on-site resources (computers, edge devices)
\end{itemize}

@@

The main contributions of this paper are i) description of a framework
for the components of the Data \& Analysis Services that builds on
current best practices from the data engineering community into the
OpenHIE framework; ii) evaluation of different implementations and
design options for type 1 and 2 data sharing within an extended OpenHIE
architecture; and iii) provide open content for the MomCare
implementation to facilitate adaptation and deployment. As such, it aims
to inform future developments and implementation of open digital health
platforms in LMICs.

@@

\subsubsection{Level of opennness}\label{level-of-opennness}

TO DO: evaluate openness of OpenHIM platform

\subsubsection{Core maintenance
functions}\label{core-maintenance-functions}

TO DO: evaluate maintenance functions

\subsubsection{Downward scaleability}\label{downward-scaleability}

TO DO: evaluate downward scalability

@@

\subsubsection{Level of opennness}\label{level-of-opennness-1}

TO DO: evaluate openness of OpenHIM platform

\subsubsection{Core maintenance
functions}\label{core-maintenance-functions-1}

TO DO: evaluate maintenance functions

\subsubsection{Downward scaleability}\label{downward-scaleability-1}

TO DO: evaluate downward scalability

@@

\subsubsection{Level of opennness}\label{level-of-opennness-2}

TO DO: evaluate openness of OpenHIM platform

\subsubsection{Core maintenance
functions}\label{core-maintenance-functions-2}

TO DO: evaluate maintenance functions

\subsubsection{Downward scaleability}\label{downward-scaleability-2}

TO DO: evaluate downward scalability

@@

\begin{itemize}
\tightlist
\item
  Object of openness: what data-related resources should data platforms
  make available when opening up (e.g.~data, data products, datadriven
  insights, analytics modules)? Which user groups derive value from
  accessing data-related resources from data platforms (e.g.~data
  providers, data users, intermediaries, developers)?
\item
  Unit of analysis: what is platform-to-platform openness in the context
  of data platforms, given the expectation that different HDPs will
  emerge at various aggregation levels? How do we distinguish
  meta-platforms, forking, and platform interoperability?
\item
  Risk of openness: What are the novel (negative) implications of
  opening up data platforms? How can reflexivity in design help
  providers to resolve the negative implications of openness?
\item
  Answers/insights to above:
